\documentclass[12pt]{article}

\setlength{\parindent}{0em}
\setlength{\parskip}{.5em}

\usepackage{framed}
\newcounter{problem}
\newcounter{problempart}[problem]
\newcounter{solutionpart}[problem]
\newenvironment{problem}{\stepcounter{problem}\noindent{\bf\arabic{problem}.}}{\setcounter{problempart}{0}\setcounter{solutionpart}{0}}
\newenvironment{solution}{\par\textcolor{green!50!black}\bgroup}{\egroup\par}
\newcommand{\qpart}{\stepcounter{problempart}${}$\\\noindent{(\alph{problempart})} }
\newcommand{\spart}{\stepcounter{solutionpart}${}$\\\noindent{(\alph{solutionpart})} }
\newcommand{\TODO}{\textcolor{red}{$\blacksquare$}}
\newcommand{\SOL}[1]{\textcolor{green!50!black}{#1}}

\usepackage{hyperref}
\usepackage{fullpage}
\usepackage{amsmath,mathabx,MnSymbol}
\usepackage{color,tikz}
\usepackage{pstricks}
\usepackage{pst-plot,pst-node}
\usepackage{footnote,enumitem}
\usepackage{longtable}
\newcommand{\mx}[1]{\begin{pmatrix}#1\end{pmatrix}}
\definecolor{dkgreen}{rgb}{0,.5,0}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

\newcommand{\uu}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\cc}{\mathbf{c}}
\newcommand{\ww}{\mathbf{w}}
\newcommand{\xx}{\mathbf{x}}
\newcommand{\zz}{\mathbf{z}}
\newcommand{\ee}{\mathbf{e}}
\newcommand{\pp}{\mathbf{p}}
\newcommand{\qq}{\mathbf{q}}
\renewcommand{\AA}{\mathbf{A}}
\newcommand{\BB}{\mathbf{B}}
\newcommand{\nn}{\mathbf{n}}
\newcommand{\gp}[1]{\left(#1\right)}

\newcommand{\TODOL}[1]{\textcolor{red}{\underline{\hspace{#1 cm}}}}

\usepackage{listings}

\lstset{
  language=C++,
  showstringspaces=false,
  identifierstyle=\color{magenta},
  basicstyle=\color{magenta},
  keywordstyle=\color{blue},
  identifierstyle=\color{black},
  commentstyle=\color{green},
  stringstyle=\color{red}
}

\begin{document}

\title{CS130 - LAB - Debugging}
\date{}
\author{Name: \TODOL7\qquad\qquad SID: \TODOL4}
\maketitle
\begin{center}
\end{center}

\newcommand{\CM}[1]{\textcolor{blue}{\texttt{#1}}}

Today's lab will be about debugging programs using GDB and valgrind. If you are
using Linux/MacOS, GDB should be already installed. You can install valgrind on
Ubuntu using \$ sudo apt get install valgrind and on MacOS using brew install
valgrind. If you are using a Windows machine, you will need to use of the
machines with Linux in the department to finish this lab.  You can connect
remotely to the machines in the lab if you want to test this at home.  There is
also some possibility of valgrind working under WSL.

\section{GDB}

GDB helps you understand the execution of the program by allowing you to run a
code line by line and check variable values on-the-fly. Say we have the
following program called factorial in a file \texttt{factorial.cpp}:

\begin{lstlisting}
float factorial(int n)
{
  float i = n;
  for(n--; n >= 0; --n)
    i *= n;
  return i;
}
\end{lstlisting}

To run GDB or valgrind, we need to compile factorial with debug symbols and we
can do this by passing \texttt{-g} to the GCC compiler.  To run factorial with
GDB, we type \CM{gdb factorial}.  This will start GDB and load the debug symbols. We
can run the program by typing \CM{run}. If the program crashes, it will stop at the
part of the code where the problem happened. To see the code where the problem
happened, you can type \CM{list}. We can also see what the value of the variables are
by typing \CM{print <variable>}.  For instance, if we want to see the value
of \texttt{n} in line 4 of factorial, we can type \CM{print n}.

Before you run the program (while in gdb), you can also add breakpoints that will make the
program stop at a specific line of code before continuing.  To do this, you can
type \CM{breakpoint <filename>:<line of code>}. For instance, if we
want to check the values \texttt{n} in \texttt{factorial.cpp}, we can type
\CM{breakpoint factorial.cpp:4}.

A quick guide to GDB can be found at:\\
\texttt{https://web.eecs.umich.edu/\textasciitilde{}sugih/pointers/summary.html}

\section*{Valgrind}

Valgrind helps us understand if there are memory violations in our program
(among other things). For instance, the following program may not crash but we
know it is wrong because we should not be accessing a memory position at index 2
of the array.

\begin{lstlisting}
int main()
{
  int *array = new int[2];
  array[0] = 0;
  array[1] = 0;
  array[2] = 0; // what will happen here?
  return array[2];
}
\end{lstlisting}

We can run valgrind by typing \CM{valgrind <program call>}. Assuming the
above code binary is called test, then we can do \CM{valgrind test}. Here is the
output of valgrind when we run test:

\begin{verbatim}
==16319== Memcheck, a memory error detector
==16319== Copyright (C) 2002-2017, and GNU GPL'd, by Julian Seward et al.
==16319== Using Valgrind-3.13.0 and LibVEX; rerun with -h for copyright info
==16319== Command: ./test
==16319== 
==16319== Invalid write of size 4
==16319==    at 0x1086B0: main (main.cpp:5)
==16319==  Address 0x5b82c88 is 0 bytes after a block of size 8 alloc'd
==16319==    at 0x4C3089F: operator new[](unsigned long) (in
/usr/lib/valgrind/vgpreload_memcheck-amd64-linux.so)
==16319==    by 0x10868B: main (main.cpp:2)
==16319== 
==16319== Invalid read of size 4
==16319==    at 0x1086BA: main (main.cpp:6)
==16319==  Address 0x5b82c88 is 0 bytes after a block of size 8 alloc'd
==16319==    at 0x4C3089F: operator new[](unsigned long) (in
/usr/lib/valgrind/vgpreload_memcheck-amd64-linux.so)
==16319==    by 0x10868B: main (main.cpp:2)
==16319== 
==16319== 
==16319== HEAP SUMMARY:
==16319==     in use at exit: 8 bytes in 1 blocks
==16319==   total heap usage: 2 allocs, 1 frees, 72,712 bytes allocated
==16319== 
==16319== LEAK SUMMARY:
==16319==    definitely lost: 8 bytes in 1 blocks
==16319==    indirectly lost: 0 bytes in 0 blocks
==16319==      possibly lost: 0 bytes in 0 blocks
==16319==    still reachable: 0 bytes in 0 blocks
==16319==         suppressed: 0 bytes in 0 blocks
==16319== Rerun with --leak-check=full to see details of leaked memory
==16319== 
==16319== For counts of detected and suppressed errors, rerun with: -v
==16319== ERROR SUMMARY: 2 errors from 2 contexts (suppressed: 0 from 0)
\end{verbatim}

The first problem is an invalid write of size 4 (bytes) on line 5. The second is
a read of the same memory address. Note also the ``definitely lost'' line which
is saying that we allocated 8 bytes but we never freed that memory, which is also
a problem.  The code we will be working with can be found on iLearn.

\begin{enumerate}
\item Run valgrind on \texttt{prog-1}.
\begin{enumerate}
\item What type of error do we get and why?
  \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
  \end{solution}
\item How can \texttt{prog-1} be changed so we don't get this error anymore?
  Note that a correct solution will allow the loop to run to 1000000 in a
  reasonable amount of time.
  \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
  \end{solution}
\end{enumerate}
Continue these two steps and update the appropriate section until there are no more errors found.

\item Run valgrind on \texttt{prog-2}.
\begin{enumerate}
\item What type of error do we get and why?
  \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
  \end{solution}
\item How \texttt{prog-2} can be changed so we don't get this error anymore?
  \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
  \end{solution}
\end{enumerate}
\item Run gdb on \texttt{prog-3}.
\begin{enumerate}
\item The program should stop with segmentation fault exception. Type list to
  see the region where the program stopped. In which line of code is the program
  crashing?
  \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
  \end{solution}
\item Use the command \CM{print <statement>} with the variables that are being
  accessed on the line where the program is crashing. You can use the \CM{up}
  command to go back a line if needed. Why does the program crash in this case
  and how we can fix it?
  \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
  \end{solution}
\end{enumerate}
\item Run valgrind on \texttt{prog-4}.
  \begin{enumerate}
  \item What type of error do we get and why?
    \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
    \end{solution}
  \item How can \texttt{prog-4} be changed so we don't get this error anymore?
    \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
    \end{solution}
  \end{enumerate}
\item Run gdb on \texttt{prog-5} and follow the steps below.
  \begin{enumerate}
  \item The program should stop with a segmentation fault exception. In which line
    of code is the program crashing?
    \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
    \end{solution}
  \item Why does the program crash in this case and how we can fix it? (You may
    want to see the list and node structures in the source code for this.)
    \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
    \end{solution}
  \item Compile and run the program again using gdb. The program should crash
    again. Try using list and print to figure out why the program is crashing and
    briefly explain your reasoning. What changes need to be made in the code to
    fix this problem?
    \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
    \end{solution}
  \item Compile and run the program again using valgrind. The program should
    display an error. Why do we get this error and how we can fix it?
    \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
    \end{solution}
  \end{enumerate}
\item Using gdb and valgrind (use the best for each situation), briefly describe
  all problems in \texttt{prog-6} and propose fixes for each one of the
  problems.
  \begin{solution}
  \textbf{\textcolor{red}{\TODO}}
  \end{solution}
\end{enumerate}

\end{document}
